---
title: "odin"
author: "Rich FitzJohn"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{odin}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

``` {r setup, echo = FALSE, results = "hide"}
lang_output <- function(x, lang) {
  cat(c(sprintf("```%s", lang), x, "```"), sep = "\n")
}
c_output <- function(x) lang_output(x, "cc")
r_output <- function(x) lang_output(x, "r")
plain_output <- function(x) lang_output(x, "plain")
knitr::opts_chunk$set(
  fig.width = 7,
  fig.height = 5)
options(odin.verbose = FALSE)
```

# Single variable: Logistic growth

We'll first start with a very simple model; logistic growth.  We
have one variable `N` which grows towards a carrying capacity `K`.
The growth rate is `r` when `N` is very small (note that this
equation can be easily solved analytically and it's only included
here for demonstration!)

``` {r }
path_logistic <- system.file("examples/logistic.R", package = "odin")
```

``` {r echo = FALSE, results = "asis"}
r_output(readLines(path_logistic))
```

The time derivatives are indicated by `deriv(N) <- `; the right
hand side of this is the differential equation.  Initial conditions
are always given (future versions may make them optional) and can
be in terms of variables.  Note that the definition of declaration
is not important; the derivative calculation references parameters
`r` and `K` which have not been defined and the initial conditions
reference `N0`.

**NOTE** while this _looks_ like R (and must *parse* as R) it is
not actually R code.  Copying and pasting this code into an R
session would throw an error

To compile this as a standalone model (not suitable for inclusion
in a package) use `odin::odin`;
``` {r }
generator <- odin::odin(path_logistic)
```

This returns an R6 generator that can be used to create an instance of the model:
``` {r }
mod <- generator$new()
mod
```

This is an `ode_system` object and can be used to run the system of
differential equations, query the internal state, and so on.  Most
of the listed elements above do not need to be accessed ever (I
would make them private but that incurs a very slight performance
cost).

The initial conditions:
``` {r }
mod$init
```

dN/dt evaluated at the time zero, with the initial conditions:
``` {r }
mod$deriv(0, mod$initial(0))
```

or with arbitrary conditions:
``` {r }
mod$deriv(0, 50)
```

To see the contents of all the variables, intermediates and
parameters tracked by odin use `contents()`:
``` {r }
mod$contents()
```

To run the model, provide a set of times:
``` {r }
tt <- seq(0, 30, length.out = 101)
y <- mod$run(tt)
plot(y, xlab = "Time", ylab = "N", las = 1, main = "")
```

While initial conditions are computed, you may pass in arbitrary
initial conditions;
``` {r }
y2 <- mod$run(tt, 50)
plot(y, xlab = "Time", ylab = "N", las = 1, main = "")
lines(y2, col = "red")
```

## Specifying parameters

The above is about the simplest possible useful model I could come
up with.  But it's not that useful because all the parameters are
hard coded into the model.  Here's a slightly tweaked version where
the parameters may be specified at runtime:
``` {r }
generator <- odin::odin({
  deriv(N) <- r * N * (1 - N / K)
  initial(N) <- N0

  N0 <- user(1)
  K <- user(100)
  r <- user()
})
```

when used as `user(1)` (as for `N` and `K`) it means we allow a
user parameter called `N0` to be specified, but if omitted it has a
default value of 1.  When used as `user()` (as for `r`) it means
that there is no default value and `r` **must** be provided.

Note this example takes the code of the model as its first
argument.  It can also be specified as text, but the form above
should be nicer to write than text, especially in editors with
support for things like autocompletion.

Because `r` is required, it is an error to initialise the model
without any parameters:

``` {r error = TRUE}
generator$new()
```

Providing `r`:
``` {r }
mod <- generator$new(r = 1)
```

The model contains the set value of r
``` {r }
mod$contents()$r
```

Running the model shows the effect of doubling the growth rate:
``` {r }
y3 <- mod$run(tt)
plot(y, xlab = "Time", ylab = "N", las = 1, main = "")
lines(y3, col = "red")
```

Once created a model can have the parameters re-set:
``` {r }
mod$set_user(r = 0.25, K = 75, N0 = 10)
y4 <- mod$run(tt)
plot(y, xlab = "Time", ylab = "N", las = 1, main = "")
lines(y3, col = "red")
lines(y4, col = "blue")
```

# More than one variable: the Lorenz attractor

Once more than one variable is involved, there is a bit more
book-keeping to take care of.
``` {r }
path_lorenz <- system.file("examples/lorenz.R", package = "odin")
```

The Lorenz attractor is a set of coupled ODEs that displays chaotic
behaviour.  It was found by reducing a set of equations describing
atmospheric convection in the 1960s.  Mostly I just think it looks
pretty.

``` {r echo = FALSE, results = "asis"}
r_output(readLines(path_lorenz))
```

The system is not chaotic with all parameters.  The initial
conditions are fairly arbitrary here as the system will settle into
its characteristic shape from most points (though with completely
different realisations).

Building the generator, and from that a system, is the same as the
above:
``` {r }
generator <- odin::odin(path_lorenz)
mod <- generator$new()
```

Because of the rapid changes that characterise this model, we'll
take a _lot_ of samples.
``` {r }
tt <- seq(0, 100, length.out = 20000)
system.time(y <- mod$run(tt))
pairs(y[, -1L], panel = lines, lwd = .2, col = "#00000055")
```

# Delay models

Models with lags and delays represent a challenge in solving and
representing the system, see [this wikipedia
article](https://en.wikipedia.org/wiki/Delay_differential_equation)
for some of the issues.  In the general case this gives a set of
partial differential equations, which are much harder to solve
generally than ordinary differential equations.

Like Berkeley Madonna, `odin` allows solving equations with lags,
using the machinery available in `deSolve`.  One can specify a lag
in a state variable or an expression.  The lag times can also be
variables.
``` {r }
gen <- odin::odin({
  ylag <- delay(y, tau)
  initial(y) <- 0.5
  deriv(y) <- 0.2 * ylag * 1 / (1 + ylag^10) - 0.1 * y
  tau <- user(10)
  output(ylag) <- ylag
})
dde <- gen$new()
```

The first line here

```r
ylag <- delay(y, tau)
```

specifies the delay; it says that `ylag` should be equal to `y`
from `tau` ago.  Or; `ylag(t) = y(t - tau)`.

`delay` expressions must be the only expressions on the rhs of an
assignment.  The first argument can be a single variable (like
here) or it can be an expression.

Running the model will automatically use the `deSolve::dede`
function.
``` {r }
t <- seq(0, 300, length.out = 301)
y1 <- dde$run(t)
plot(y1, ylab = "y", mfrow = c(1, 2), which = 1)
plot(y1[, -1L], xlab = "y", ylab = "ylag", mfrow = NULL, type = "l")
```

Increasing the lag time creates much more complicated trajectories:
``` {r }
dde$set_user(tau = 20)
y2 <- dde$run(t)
plot(y2, ylab = "y", mfrow = c(1, 2), which = 1)
plot(y2[, -1L], xlab = "y", ylab = "ylag", mfrow = NULL, type = "l")
```

# Arrays

Much of the above, especially the non-delay equations, can be
expressed easily in deSolve.  Even where the target functions are
written in R they will tend to be fast enough to solve (unless
strongly non-linear) so that the gains for writing out in C become
small.  `odin` was really designed for cases where the variables
involved are _arrays_, used here to mean vectors, matrices or 3d
matrices.

There are quite a few restrictions on how arrays are used, how
indexing works, etc.  Some of the current implementation requires
creation of intermediate variables to make everything work (and
naming things is one of the [two hard problems in computer
science](https://martinfowler.com/bliki/TwoHardThings.html).
Hopefully in future versions some of these restrictions can be
relaxed, so [file issues](https://github.com/mrc-ide/odin/issues)
if you find yourself writing boilerplate code that could be
replaced by more intelligent generation.

## Generalised Lotka-Volterra model

This example comes from the [wikipedia
page](https://en.wikipedia.org/wiki/Competitive_Lotka%E2%80%93Volterra_equations#4-dimensional_example),
in turn from [Vano et al
2006](https://sprott.physics.wisc.edu/pubs/paper288.pdf).  This
model has four non-linear differential equations that have chaotic
behaviour in fairly low dimension.
``` {r }
gen <- odin::odin({
  deriv(y[]) <- r[i] * y[i] * (1 - sum(ay[i, ]))
  initial(y[]) <- y0[i]

  y0[] <- user()
  r[] <- user()
  a[, ] <- user()
  ay[, ] <- a[i, j] * y[j]

  dim(r) <- user()
  n_spp <- length(r)

  dim(y) <- n_spp
  dim(y0) <- n_spp
  dim(a) <- c(n_spp, n_spp)
  dim(ay) <- c(n_spp, n_spp)

  config(base) <- "lv4"
})
```

These are the parameters from the wikipedia article, with `y0` being
close to the (unstable) equilibrium.
``` {r }
pars <- list(r = c(1.00, 0.72, 1.53, 1.27),
             a = rbind(c(1.00, 1.09, 1.52, 0.00),
                       c(0.00, 1.00, 0.44, 1.36),
                       c(2.33, 0.00, 1.00, 0.47),
                       c(1.21, 0.51, 0.35, 1.00)),
             y0 = c(0.3013, 0.4586, 0.1307, 0.3557))
```

Note that we pass in `r` as a matrix, from which the number of
species is determined.  The competitive matrix `a` is passed
through as a matrix -- both dimensions must be the same as the
length of `r` (you can't just pass in a vector and hope it gets
unpacked in the right order).
``` {r }
mod <- gen$new(user = pars)

t <- seq(0, 2000, length.out = 10001)
y <- mod$run(t)
pairs(y[, -1], panel = lines, col = "#00000055", lwd = 0.2)
```

# Interpolating functions

It may be useful to have functions that are driven by user data in
your model.  To allow this, odin allows interpolating functions of
a few different types;

* step functions: constant piecewise functions, perhaps suitable
  for modelling shifts in regime, binary on/off switches, etc.
* linear interpolation: between a set of points use a linear
  interpolating function.
* cubic spline functions: what most people probably think of when
  thinking interpolating function; creates a smooth function that
  is twice differentiable (and smooth enough for most purposes).
  In contrast with the splines R uses with `spline()`, these use
  the *natural spline* end conditions, and are equal to `spline(t,
  y, method = "natural")`.

The `interpolate` takes as its first argument a vector of times.
If you're using constant or linear interpolating functions it is
important that you include these times within times that the
integrator passes through because otherwise the integrator may
struggle needlessly with the discontinuities in the function or its
derivatives (NOTE: future versions may handle this automatically,
and this will not work at all with `dde` which does not yet support
stopping at critical times).

The second argument of each interpolating function is a vector,
matrix or array of values for the target function.  The
dimensionality (rank) of the structure must be one greater than the
rhs.  So if you want a classic 1d function, you'd write:

```r
z <- interpolate(tt, y)
tt[] <- user()
y[] <- user()
dim(tt) <- user()
dim(y) <- length(tt)
```

and `z` will be a scalar double, and `t` and `y` are both vectors
with lengths that must match.  The check for matching is done at
runtime (when the model is initialised) so you can always just
leave the dimensions as `user()`.  Future versions may allow
autogeneration of the boilerplate here, but for now you must
specify the four lines required to initialise `t` and `y`.

Note that `t` cannot be used as
the first argument because it's a reserved variable!

If you want to interpolate a *vector* of `y` values over time, then
you'd write:

```r
z[] <- interpolate(tt, y)
dim(z) <- 4
tt[] <- user()
y[, ] <- user()
dim(tt) <- user()
dim(y) <- c(length(tt), length(z))
```

In this case, `z` is a vector (here length 4), so `y` must be a
matrix with `length(t)` rows and 4 columns.  This does not do
anything clever - the interpolation happens for each of the four
variables independently.

If you want to drive the size of this off of `y` then specify:

```r
dim(z) <- ncol(y)
dim(z) <- user()
# (other entries same as above)
```

in which case the size of `z` is determined from the number of
columns of `y`.

By default, `odin` assumes you want do do cubic interpolation, but
you can alter that with the third argument.  Specify `"constant"`
to indicate constant interpolation and `"linear"` to indicate
linear interpolation.

All modes of interpolation use a fast lookup (bisection search
that remembers where it started from on the last lookup) to try and
make interpolation as fast as possible.  Performance should be
*O(log(n))* in the number of points used to create the
interpolating function, so the penalty for using many points in
your interpolation should be low.

No care is taken about making sure that the integration stops at
painful parts of your interpolating function.  So, especially for
the constant interpolation, be sure to include the switch points in
the time vector (currently this will not work for `dde` which never
stops at a specified point except for the start and end point).
Future versions, especially once events are handled, may do this
automatically.

As an example, here is the flux model from the deSolve compiledCode
vignette (see `vignette("compiledCode")`, section 6).

The model here is:

dC/dt = flux(t) - k * C
``` {r }
flux_model <- odin::odin({
  deriv(C) <- flux - kk * C
  initial(C) <- C0
  flux <- interpolate(flux_t, flux_y, "linear")
  C0 <- user()
  kk <- user()
  output(deposition) <- kk * C
  ## Fair bit of boilerplate here that may be removed in future
  ## versions:
  flux_t[] <- user()
  flux_y[] <- user()
  dim(flux_t) <- user()
  dim(flux_y) <- user()
})
```

Here we arrange for the components of the interpolating function to
be passed through as user vectors, which is the only way of
specifying these models at present.  The interpolation type, here
"linear", is hard coded and cannot be changed at runtime.
``` {r }
flux_t <- c(1, 11, 21, 41, 73, 83, 93, 103, 113, 123, 133, 143, 153,
            163, 173, 183, 194, 204, 214, 224, 234, 244, 254, 264,
            274, 284, 294, 304, 315, 325, 335, 345, 355, 365)
flux_y <- c(0.654, 0.167, 0.06, 0.07, 0.277, 0.186, 0.14, 0.255, 0.231,
            0.309, 1.127, 1.923, 1.091, 1.001, 1.691, 1.404, 1.226, 0.767,
            0.893, 0.737, 0.772, 0.726, 0.624, 0.439, 0.168, 0.28, 0.202,
            0.193, 0.286, 0.599, 1.889, 0.996, 0.681, 1.135)
plot(flux_t, flux_y, type = "l", ylab = "Flux", xlab = "Time")
```

Following the deSolve vignette, the initial conditions are derived
from the parameters and the flux over time:
``` {r }
k <- 0.01
C0 <- mean(approx(flux_t, flux_y, xout = 1:365)$y) / k
```

Initialise the model:
``` {r }
mod <- flux_model$new(kk = k, C0 = C0, flux_t = flux_t, flux_y = flux_y)
```

And run over one year:
``` {r }
t <- seq(1, 365)
y <- mod$run(t, tcrit = 365)
```

The output in all its glory:
``` {r }
plot(flux_t, flux_y, type = "l", col = "red", ylab = "Flux & deposition")
lines(t, y[, 3], col = "blue")
```

Alternatively, we could have used constant or spline interpolation.
This (at least at present) requires re-specifying and recompiling
the entire model.  So, for a constant model:
``` {r }
flux_model_c <- odin::odin({
  deriv(C) <- flux - kk * C
  initial(C) <- C0
  flux <- interpolate(flux_t, flux_y, "constant")
  C0 <- user()
  kk <- user()
  output(flux) <- flux
  output(deposition) <- kk * C
  ## Fair bit of boilerplate here that may be removed in future
  ## versions:
  flux_t[] <- user()
  flux_y[] <- user()
  dim(flux_t) <- user()
  dim(flux_y) <- user()
})

mod_c <- flux_model_c$new(kk = k, C0 = C0, flux_t = flux_t, flux_y = flux_y)
y_c <- mod_c$run(t, tcrit = 365)

mod_c$output_order
plot(t, y_c[, 3], type = "s", col = "red", ylab = "Flux & deposition")
lines(t, y_c[, 4], col = "blue")
```

Or with cubic splines:
``` {r }
flux_model_s <- odin::odin({
  deriv(C) <- flux - kk * C
  initial(C) <- C0
  flux <- interpolate(flux_t, flux_y, "spline")
  C0 <- user()
  kk <- user()
  output(flux) <- flux
  output(deposition) <- kk * C
  ## Fair bit of boilerplate here that may be removed in future
  ## versions:
  flux_t[] <- user()
  flux_y[] <- user()
  dim(flux_t) <- user()
  dim(flux_y) <- user()
})

mod_s <- flux_model_s$new(kk = k, C0 = C0, flux_t = flux_t, flux_y = flux_y)
y_s <- mod_s$run(t, tcrit = 365)
plot(t, y_s[, 3], type = "l", col = "red", ylab = "Flux & deposition")
lines(t, y_s[, 4], col = "blue")
```
