---
title: "Odin"
author: "Rich FitzJohn"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Odin}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

``` {r setup, echo=FALSE, results="hide"}
lang_output <- function(x, lang) {
  cat(c(sprintf("```%s", lang), x, "```"), sep="\n")
}
c_output <- function(x) lang_output(x, "cc")
r_output <- function(x) lang_output(x, "r")
plain_output <- function(x) lang_output(x, "plain")
knitr::opts_chunk$set(
  fig.width=7,
  fig.height=5)
```

# Single variable: Logistic growth

We'll first start with a very simple model; logisitic growth.  We
have one variable `N` which grows towards a carrying capacity `K`.
The growth rate is `r` when `N` is very small (note that this
equation can be easily solved analytically and it's only included
here for demonstration!)

``` {r }
path_logistic <- system.file("examples/logistic.R", package="odin")
```

``` {r echo=FALSE, results="asis"}
r_output(readLines(path_logistic))
```

The time derivatives are indicated by `deriv(N) <- `; the right
hand side of this is the differential equation.  Initial conditions
are always given (future versions may make them optional) and can
be in terms of variables.  Note that the definition of declaration
is not important; the derivative calculation references parameters
`r` and `K` which have not been defined and the initial conditions
reference `N0`.

**NOTE** while this _looks_ like R (and must *parse* as R) it is
not actually R code.  Copying and pasting this code into an R
session would throw an error

To compile this as a standalone model (not suitable for inclusion
in a package) use `odin::odin`;
``` {r }
generator <- odin::odin(path_logistic, verbose=FALSE)
```

This returns a function that will generate an instance of the model:
``` {r }
mod <- generator()
mod
```

This is an `ode_system` object and can be used to run the system of
differential equations, query the internal state, and so on.  Most
of the listed elements above do not need to be accessed ever (I
would make them private but that incurrs a very slight performance
cost).

The initial conditions:
``` {r }
mod$init
```

dN/dt evaluated at the time zero, with the initial conditions:
``` {r }
mod$deriv(0, mod$init)
```

or with arbitrary conditions:
``` {r }
mod$deriv(0, 50)
```

To see the contents of all the variables, intermediates and
parameters tracked by odin use `contents()`:
``` {r }
mod$contents()
```

To run the model, provide a set of times:
``` {r }
tt <- seq(0, 30, length.out=101)
y <- mod$run(tt)
plot(y, xlab="Time", ylab="N", las=1, main="")
```

While initial conditions are computed, you may pass in arbitrary
initial conditions;
``` {r }
y2 <- mod$run(tt, 50)
plot(y, xlab="Time", ylab="N", las=1, main="")
lines(y2, col="red")
```

## Specifying parameters

The above is about the simplest possible useful model I could come
up with.  But it's not that useful because all the parameters are
hard coded into the model.  Here's a slightly tweaked version where
the parameters may be specified at runtime:
``` {r }
generator <- odin::odin({
  deriv(N) <- r * N * (1 - N / K)
  initial(N) <- N0

  N0 <- user(1)
  K <- user(100)
  r <- user()
}, verbose=FALSE)
```

when used as `user(1)` (as for `N` and `K`) it means we allow a
user parameter called `N0` to be specified, but if omitted it has a
default value of 1.  When used as `user()` (as for `r`) it means
that there is no default value and `r` **must** be provided.

Note this example takes the code of the model as its first
argument.  It can also be specified as text, but the form above
should be nicer to write than text, especially in editors with
support for things like autocompletion.

Because `r` is required, it is an error to initialise the model
without any parameters:

``` {r error=TRUE}
generator()
```

Providing `r`:
``` {r }
mod <- generator(r=1)
```

The model contains the set value of r
``` {r }
mod$contents()$r
```

Running the model shows the effect of doubling the growth rate:
``` {r }
y3 <- mod$run(tt)
plot(y, xlab="Time", ylab="N", las=1, main="")
lines(y3, col="red")
```

Once created a model can have the parameters re-set:
``` {r }
mod$set_user(r=0.25, K=75, N0=10)
y4 <- mod$run(tt)
plot(y, xlab="Time", ylab="N", las=1, main="")
lines(y3, col="red")
lines(y4, col="blue")
```

Note that it is not (currently) an error to set unknown parameters.
This will be fixed in a future version:
``` {r }
mod$set_user(whatever=1)
```

# More than one variable: the Lorenz attractor

Once more than one variable is involved, there is a bit more
book-keeping to take care of.
``` {r }
path_lorenz <- system.file("examples/lorenz.R", package="odin")
```

The Lorenz attractor is a set of coupled ODEs that displays chaotic
behaviour.  It was found by reducing a set of equations describing
atmospheric convection in the 1960s.  Mostly I just think it looks
pretty.

``` {r echo=FALSE, results="asis"}
r_output(readLines(path_lorenz))
```

The system is not chaotic with all parameters.  The initial
conditions are fairly arbitrary here as the system will settle into
its characteristic shape from most points (though with completely
different realisations).

Building the generator, and from that a system, is the same as the
above:
``` {r }
generator <- odin::odin(path_lorenz, verbose=FALSE)
mod <- generator()
```

Because of the rapid changes that characterise this model, we'll
take a _lot_ of samples.
``` {r }
tt <- seq(0, 100, length.out=20000)
system.time(y <- mod$run(tt))
pairs(y[, -1L], panel=lines, lwd=.2, col="#00000055")
```

# Delay models

Models with lags and delays represent a challenge in solving and
representing the system see [this wikipedia
article](https://en.wikipedia.org/wiki/Delay_differential_equation)
for some of the issues.  In the general case this gives a set of
partial differential equations, which are much harder to solve
generally than ordinary differential equations.

Like Berkeley Madonna, `odin` allows solving equations with lags,
using the machinery available in `deSolve`.  One can specify a lag
in a state variable or an expression.  The lag times can also be
variables.

**NOTE** the current version supports only scalar delays; delays
that involve arrays are not currently handled, but will be soon.
``` {r }
gen <- odin::odin({
  ylag <- delay(y, tau)
  initial(y) <- 0.5
  deriv(y) <- 0.2 * ylag * 1 / (1 + ylag^10) - 0.1 * y
  tau <- user(10)
  output(ylag) <- ylag
})
dde <- gen()
```

The first line here

```r
ylag <- delay(y, tau)
```

Specifies the delay; it says that `ylag` should be equal to `y`
from `tau` ago.  Or; `ylag(t) = y(t - tau)`.

`delay` expressions must be the only expressions on the rhs of an
assignment.  The first argument can be a single variable (like
here) or it can be an expression.

Running the model will automatically use the `deSolve::dede`
function.
``` {r }
t <- seq(0, 300, length.out=301)
y1 <- dde$run(t)
plot(y1, ylab = "y", mfrow=c(1, 2), which=1)
plot(y1[, -1L], xlab="y", ylab = "ylag", mfrow=NULL, type="l")
```

Increasing the lag time creates much more complicated trajectories:
``` {r }
dde$set_user(tau=20)
y2 <- dde$run(t)
plot(y2, ylab = "y", mfrow=c(1, 2), which=1)
plot(y2[, -1L], xlab="y", ylab = "ylag", mfrow=NULL, type="l")
```

# Arrays

Much of the above, especially the non-delay equations, can be
expressed easily in deSolve.  Even where the target functions are
written in R they will tend to be fast enough to solve (unless
strongly nonlinear) so that the gains for writing out in C become
small.  `odin` was really designed for cases where the variables
involved are _arrays_, used here to mean vectors, matrices or 3d
matrices.

There are quite a few restrictions on how arrays are used, how
indexing works, etc.  Some of the current implementation requires
creation ofintermediate variables to make everything work (and
naming things is one of the [two hard problems in computer
science](https://www.quora.com/Why-is-naming-things-hard-in-computer-science-and-how-can-it-can-be-made-easier).
Hopefully in future versions some of these restrictions can be
relaxed, so [file issues](https://github.com/richfitz/odin/issues)
if you find yourself writing boilerplate code that could be
replaced by more intelligent generation.

## Generalised Lotka-Volterra model

This example comes from the [wikipedia
page](https://en.wikipedia.org/wiki/Competitive_Lotka%E2%80%93Volterra_equations#4-dimensional_example),
in turn from [Vano et al
2006](http://sprott.physics.wisc.edu/pubs/paper288.pdf).  This
model has four nonlinear differential equations that have chaotic
behaviour in fairly low dimension.
``` {r }
gen <- odin::odin({
  deriv(y[]) <- r[i] * y[i] * (1 - sum(ay[i, ]))
  initial(y[]) <- y0[i]

  y0[] <- user()
  r[] <- user()
  a[,] <- user()
  ay[,] <- a[i, j] * y[j]

  dim(r) <- user()
  n_spp <- length(r)

  dim(y) <- n_spp
  dim(y0) <- n_spp
  dim(a) <- c(n_spp, n_spp)
  dim(ay) <- c(n_spp, n_spp)

  config(base) <- "lv4"
})
```

These are the parameters from the wikipedia article, with `y0` being
close to the (unstable) equilibrium.
``` {r }
pars <- list(r=c(1.00, 0.72, 1.53, 1.27),
             a=rbind(c(1.00, 1.09, 1.52, 0.00),
                     c(0.00, 1.00, 0.44, 1.36),
                     c(2.33, 0.00, 1.00, 0.47),
                     c(1.21, 0.51, 0.35, 1.00)),
             y0=c(0.3013, 0.4586, 0.1307, 0.3557))
```

Note that we pass in `r` as a matrix, from which the number of
species is determined.  The competitive matrix `a` is passed
through as a matrix -- both dimensions must be the same as the
length of `r` (you can't just pass in a vector and hope it gets
unpacked in the right order).
``` {r }
mod <- gen(user=pars)

t <- seq(0, 2000, length.out=10001)
y <- mod$run(t)
pairs(y[, -1], panel=lines, col="#00000055", lwd=0.2)
```
