---
title: "Odin"
author: "Rich FitzJohn"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Odin}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

``` {r setup, echo=FALSE, results="hide"}
lang_output <- function(x, lang) {
  cat(c(sprintf("```%s", lang), x, "```"), sep="\n")
}
c_output <- function(x) lang_output(x, "cc")
r_output <- function(x) lang_output(x, "r")
plain_output <- function(x) lang_output(x, "plain")
knitr::opts_chunk$set(
  fig.width=7,
  fig.height=5)
```

# Single variable: Logistic growth

We'll first start with a very simple model; logisitic growth.  We
have one variable `N` which grows towards a carrying capacity `K`.
The growth rate is `r` when `N` is very small (note that this
equation can be easily solved analytically and it's only included
here for demonstration!)

``` {r }
path_logistic <- system.file("examples/logistic.R", package="odin")
```

``` {r echo=FALSE, results="asis"}
r_output(readLines(path_logistic))
```

The time derivatives are indicated by `deriv(N) <- `; the right
hand side of this is the differential equation.  Initial conditions
are always given (future versions may make them optional) and can
be in terms of variables.  Note that the definition of declaration
is not important; the derivative calculation references parameters
`r` and `K` which have not been defined and the initial conditions
reference `N0`.

**NOTE** while this _looks_ like R (and must *parse* as R) it is
**not actually R code.  Copying and pasting this code into an R
**session would throw an error

To compile this as a standalone model (not suitable for inclusion
in a package) use `odin::odin`;
``` {r }
generator <- odin::odin(path_logistic, tempdir(), verbose=FALSE)
generator
```

This returns an `R6ClassGenerator` object which in turn will
generate an instance of the model:
``` {r }
mod <- generator$new()
mod
```

This is an `ode_system` object and can be used to run the system of
differential equations, query the internal state, and so on.  Most
of the listed elements above do not need to be accessed ever (I
would make them private but that incurrs a very slight performance
cost).

The initial conditions:
``` {r }
mod$init
```

dN/dt evaluated at the time zero, with the initial conditions:
``` {r }
mod$deriv(0, mod$init)
```

or with arbitrary conditions:
``` {r }
mod$deriv(0, 50)
```

To see the contents of all the variables, intermediates and
parameters tracked by odin use `contents()`:
``` {r }
mod$contents()
```

To run the model, provide a set of times:
``` {r }
tt <- seq(0, 30, length.out=101)
y <- mod$run(tt)
plot(y, xlab="Time", ylab="N", las=1, main="")
```

While initial conditions are computed, you may pass in arbitrary
initial conditions;
``` {r }
y2 <- mod$run(tt, 50)
plot(y, xlab="Time", ylab="N", las=1, main="")
lines(y2, col="red")
```

## Specifying parameters

The above is about the simplest possible useful model I could come
up with.  But it's not that useful because all the parameters are
hard coded into the model.  Here's a slightly tweaked version where
the parameters may be specified at runtime:
``` {r }
generator <- odin::odin({
  deriv(N) <- r * N * (1 - N / K)
  initial(N) <- N0

  N0 <- user(1)
  K <- user(100)
  r <- user()
}, tempdir(), verbose=FALSE)
```


when used as `user(1)` (as for `N` and `K`) it means we allow a
user parameter called `N0` to be specified, but if omitted it has a
default value of 1.  When used as `user()` (as for `r`) it means
that there is no default value and `r` **must** be provided.

Note this example takes the code of the model as its first
argument.  It can also be specified as text, but the form above
should be nicer to write than text, especially in editors with
support for things like autocompletion.

Because `r` is required, it is an error to initialise the model
without any parameters:

``` {r error=TRUE}
generator$new()
```

**NOTE:** the interface here will change so that it will be
possible to call the generator as `generator$new(r=1)` but that
needs a little work.
``` {r }
mod <- generator$new(list(r=1))
```

The model contains the set value of r
``` {r }
mod$contents()$r
```

Running the model shows the effect of doubling the growth rate:
``` {r }
y3 <- mod$run(tt)
plot(y, xlab="Time", ylab="N", las=1, main="")
lines(y3, col="red")
```

Once created a model can have the parameters re-set:
``` {r }
mod$set_user(list(r=0.25, K=75, N0=10))
y4 <- mod$run(tt)
plot(y, xlab="Time", ylab="N", las=1, main="")
lines(y3, col="red")
lines(y4, col="blue")
```

Note that it is not (currently) an error to set unknown parameters.
This will be fixed in a future version:
``` {r }
mod$set_user(list(r=1, whatever=1))
```

# More than one variable: the Lorenz attractor

Once more than one variable is involved, there is a bit more
book-keeping to take care of.
``` {r }
path_lorenz <- system.file("examples/lorenz.R", package="odin")
```

The Lorenz attractor is a set of coupled ODEs that displays chaotic
behaviour.  It was found by reducing a set of equations describing
atmospheric convection in the 1960s.  Mostly I just think it looks
pretty.

``` {r echo=FALSE, results="asis"}
r_output(readLines(path_lorenz))
```

The system is not chaotic with all parameters.  The initial
conditions are fairly arbitrary here as the system will settle into
its characteristic shape from most points (though with completely
different realisations).

Building the generator, and from that a system, is the same as the
above:
``` {r }
generator <- odin::odin(path_lorenz, tempdir(), verbose=FALSE)
mod <- generator$new()
```

Because of the rapid changes that characterise this model, we'll
take a _lot_ of samples.
``` {r }
tt <- seq(0, 100, length.out=20000)
system.time(y <- mod$run(tt))
pairs(y[, -1L], panel=lines, lwd=.2, col="#00000055")
```
